Structure Learning for $\VAR(p)$ Models

### 2019-04-28 paper.tex
Proof of prop 4 and more work on algorithm proof.

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  % nth generation grandparents
\newcommand{\gcgpath}[4]{#1 \rightarrow #2_1 \rightarrow \cdots \rightarrow #2_{#3} \rightarrow #4}

@@

\begin{proposition}
  \label{lem:pwgc_anc}
  If $\gcg$ is a strongly causal DAG then $j \in \anc{i} \implies j \pwgc i$.
\end{proposition}
\begin{proof}
  Using the representation of equation (\ref{eqn:parent_expansion}) we can represent $x_i(t)$ as

  \begin{equation*}
    x_i(t) = v_i(t) + \B_{ii}(z) x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z) x_k(t).
  \end{equation*}

  In order to show that $j \pwgc i$ we can show (due to theorem (\ref{thm:granger_causality_equivalences})) that

  \begin{equation*}
    \linE{x_i(t)}{\H_{t - 1}^{(i)}} = \linE{x_i(t)}{\H_{t - 1}^{(i, j)}},
  \end{equation*}

\hl{WIP}

% this allows us to recursively expand the
%   parents of $x_i$ in $\gcg$ in the $\VAR(\infty)$ representation of
%   $x(t)$ since we will eventually terminate at nodes $d$ such that
%   $\pa{d} = \emptyset$.  To this end, examining $x_i$ in equation
%   \ref{eqn:ar_representation}

%   \begin{align*}
%     x_i(t) &= v_i(t) + \sum_{\tau = 1}^\infty \big[ B_{ii}(\tau)x_i(t - \tau) + \sum_{k \in \pa{i}} B_{ik}(\tau)x_k(t - \tau)\big]\\
%     \implies x_i(t) &= (1 - \B_{ii}(z))^{-1} \big[v_i(t) + \sum_{k \in \pa{i}} B_{ik}(\tau)x_k(t - \tau) \big]\\
%     &= \sum_{\tau = 1}^\infty \big[C_{ii}(\tau)v_i(t - \tau) + \sum_{k \in \pa{i}} C_{ik}(\tau)x_k(t - \tau)\big].
%   \end{align*}

%   If $\pa{i} = \emptyset$ then we simply have
%   $x_i(t) = \sum_{\tau = 1}^\infty C_{ii}(\tau)v_i(t)$ and
%   $C_{ii} = A_{ii}$ from equation \ref{eqn:wold}.

%   Applying this relation recursively we can write

%   \begin{equation*}
%     x_i(t) = \sum_{\tau = 1}^\infty\big[C_{ii}(\tau)v_i(t - \tau) + \sum_{k \in \anc{i}}C_{ik}(\tau)v_k(t - \tau) \big].
%   \end{equation*}

That strong conditions need to be placed on the topology of $\gcg$ in
order for proposition \ref{lem:pwgc_anc} to be true can be seen through the
earlier examples.

@@

  This follows immediately from the fact (lemma \ref{lem:pwgc_anc})
  that $j \in \anc{i} \implies j \pwgc i$.
\end{example}

@@

  Since $j \in \anc{i}$ there is a path $\gcgpath{j}{a}{p}{i}$.
  Moreover, since $i \pwgc j$ by lemma \ref{lem:pwgc_anc} there must
  be a common ancestor $k \in \anc{i} \cap \anc{j}$ (otherwise, we would
  have $i \in \anc{j}$, contradicting the acyclic property).  The node
  $k$ has paths to both $j,\ \gcgpath{k}{b}{q}{j}$ and $i,\ \gcgpath{k}{c}{r}{i}$.

@@

      \;
      $C_k \leftarrow \{i \in S_{k + 1}\ |\ \forall s \in S_{k + 1}\ s \not\pwgc i\}$   \texttt{\# potential children of }$P_k$\\

@@

@@

  Firstly, if $n = 0$ or $n = 1$ the algorithm is trivially correct, returning on the first iteration with $E = \emptyset$.  Suppose then that $n > 1$, we will proceed by induction.

On the first iteration of the algorithm ($k = 0$) we have $S_0 = [n]$ and $E = \emptyset$.  The set $\widehat{P}_0$ is non-empty since $\gcg$ is a DAG and therefore there are nodes in $\gcg$ without parents, thence if $\widehat{P}_0 = \emptyset$ proposition \ref{prop:ancestor_properties} would be contradicted.  Now, if $\widehat{P}_0 = [n]$ then the graph necessarily contains no edges and we will correctly return $E = \emptyset$.  On the other hand, if $\widehat{P}_0 \ne [n]$ then $C_0$ must also be non-empty, otherwise there would be driving nodes not included in $\widehat{P}_0$.  Consider now an edge $(i, j) \in D_0$ ($D_0$ is non-empty for the same reason as $C_0$).  Since $i$ has no parents, $\anc{i} \cap \anc{j} = \emptyset$ and therefore $i \in \anc{j}$ (proposition \ref{prop:ancestor_properties}).  Furthermore, the construction of $C_k$ implies that $i \in \pa{j}$ since otherwise, there would be some $u \in \anc{j}$ with $i \in \anc{u}$, but then by proposition \ref{lem:pwgc_anc} $i \pwgc u$ and $u \pwgc j$ so that $u \not \in \widehat{P}_0 \implies u \in S_0 \implies j \not\in C_0$, a contradiction.

>>>>

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  % nth generation grandparents
\newcommand{\gcgpath}[4]{#1 \rightarrow #2_1 \rightarrow \cdots \rightarrow #2_{#3} \rightarrow #4}
\newcommand{\wtalpha}[2]{\widetilde{\alpha}(#1, #2)}  % Some notation for lem:pwgc_anc
\newcommand{\dist}[2]{\mathsf{d}(#1, #2)}  % Distance between things

@@

\begin{proposition}
  \label{prop:pwgc_anc}
  If $\gcg$ is a strongly causal DAG then $j \in \anc{i} \implies j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that for some $\psi \in \H_{t - 1}^{(j)}$ we have

  \begin{equation}
    \label{eqn:cond_ortho_proof}
    \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}} \ne 0
  \end{equation}

  and therefore that $H_t^{(i)} \not\perp\ \H_{t - 1}^{(j)}\ |\ \H_{t - 1}^{(i)}$, which by theorem (\ref{thm:granger_causality_equivalences}) and \hl{lemma [name the lemma]} is enough to establish that $j \pwgc i$.

  Firstly, we will establish a representation of $x_i(t)$ that involves $x_j(t)$.  Denote by $a_{r + 1} \rightarrow a_r \rightarrow \cdots \rightarrow a_1 \rightarrow a_0$ with $a_{r + 1} \defeq j$ and $a_0 \defeq i$ the \textit{unique} $j \rightarrow i$ path in $\gcg$, we will expand the representation of equation (\ref{eqn:parent_expansion}) backwards along this path:

  \begin{align*}
    x_i(t) &= v_i(t) + \B_{ii}(z) x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z) x_k(t)\\
           &= \underbrace{v_{a_0}(t) + \B_{a_0a_0}(z) x_i(t) + \sum_{\substack{k \in \pa{a_0} \\ k \ne a_1}}\B_{a_0 k}(z) x_k(t)}_{\defeq \wtalpha{a_0}{a_1}} + \B_{a_0a_1}(z)x_{a_1}(t)\\
           &= \wtalpha{a_0}{a_1} + \B_{a_0a_1}(z)\big[\wtalpha{a_1}{a_2} + \B_{a_1a_2}(z)x_{a_2}(t) \big]\\
           &\overset{(a)}{=} \cdots\\
           &= \sum_{\ell = 0}^r \underbrace{\Big(\prod_{m = 0}^{\ell - 1} \B_{a_m a_{m + 1}}(z) \Big)}_{\defeq F_\ell(z)} \wtalpha{a_\ell}{a_{\ell + 1}} + \Big(\prod_{m = 0}^{r}\B_{a_m a_{m + 1}}(z)\Big)x_{a_{r + 1}}(t)\\
           &= \sum_{\ell = 0}^r F_\ell(z) \wtalpha{a_\ell}{a_{\ell + 1}} + F_{r + 1}(z) x_j(t)
  \end{align*}

  where $(a)$ follows by a routine induction argument and where we define $\prod_{m = 0}^{-1} \bullet \defeq 1$ for notational convenience.

  Using this representation to expand equation (\ref{eqn:cond_ortho_proof}), we obtain the following cumbersome expression:

  \begin{align*}
    &\inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{F_{r + 1}(z)x_j(t) - \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}}\\
    &- \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{\linE{\sum_{\ell = 0}^r F_\ell(z)\wtalpha{a_\ell}{a_{\ell + 1}}}{\H_{t - 1}^{(i)}}}\\
    &+ \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{\sum_{\ell = 0}^r F_\ell(z)\wtalpha{a_\ell}{a_{\ell + 1}}}.
  \end{align*}

  Note that by the orthogonality principle, $\psi - \linE{\psi}{\H_{t - 1}^{(i)}} \perp \H_{t - 1}^{(i)}$, the middle term above is $0$.  Choosing now the particular value $\psi = F_{r + 1}(z)x_j(t) \in \H_{t - 1}^{(j)}$ we arrive at

  \begin{align*}
    &\inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}\\
    &= \E|F_{r + 1}(z)x_j(t) - \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}|^2\\
    &+ \inner{F_{r + 1}(z)x_j(t) - \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}}}{\sum_{\ell = 0}^r F_\ell(z) \wtalpha{a_\ell}{a_{\ell + 1}}},
  \end{align*}

  which by the Cauchy-Schwarz inequality is $0$ if and only if

  \begin{equation*}
    \sum_{\ell = 0}^r F_\ell(z) \wtalpha{a_\ell}{a_{\ell + 1}} \overset{\text{a.s.}}{=} \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}} - F_{r + 1}(z)x_j(t),
  \end{equation*}

  or by rearranging and applying the representation obtained earlier, if and only if

  \begin{equation*}
    x_i(t) \overset{\text{a.s.}}{=} \linE{F_{r + 1}(z)x_j(t)}{\H_{t - 1}^{(i)}},
  \end{equation*}

  but this is impossible since with probability one $v_i(t) \not \in \H_{t - 1}^{(i)}$.
\end{proof}

\begin{corollary}
  \label{cor:gc_implies_pwgc}
  If $\gcg$ is a strongly causal DAG then $j \gc i \implies j \pwgc i$.

  \hl{This must be true in general, should include this shortly after the definition ow PWGC}
\end{corollary}

\begin{remark}
  It can be seen from the conclusion of the proof that simple facts
  about Granger causality are indeed reliant on our assumptions about
  the nature of $x(t)$ laid out in the section \ref{sec:theory}, in
  particular, the innovations process $v(t)$ must be full rank.

  Moreover, that strong conditions need to be placed on the topology
  of $\gcg$ in order for proposition \ref{prop:pwgc_anc} to hold can
  be seen through the earlier examples.
\end{remark}

@@

  This follows immediately from the fact (lemma \ref{prop:pwgc_anc})
  that $j \in \anc{i} \implies j \pwgc i$.
\end{example}

@@

  Since $j \in \anc{i}$ there is a path $\gcgpath{j}{a}{p}{i}$.
  Moreover, since $i \pwgc j$ by lemma \ref{prop:pwgc_anc} there must
  be a common ancestor $k \in \anc{i} \cap \anc{j}$ (otherwise, we would
  have $i \in \anc{j}$, contradicting the acyclic property).  The node
  $k$ has paths to both $j,\ \gcgpath{k}{b}{q}{j}$ and $i,\ \gcgpath{k}{c}{r}{i}$.

@@

      \;
      $C_k \leftarrow \{j \in S_{k + 1}\ |\ \forall s \in S_{k + 1}\ s \not\pwgc j\}$   \texttt{\# potential children of }$P_k$\\

@@

\begin{lemma}
  \hl{Some lemma on the inner loop.}
\end{lemma}
\begin{proof}
  Let $j \in C_k$ and $i \in \bigcup_{r = 0}^k P_{k - r}$.

  First, suppose that $i \in \pa{j}$ and that there is no $i \rightarrow j$ path in $E$.  Then since $i \gc j$ we have $i \pwgc j$ by proposition \ref{prop:pwgc_anc} so at least one of $\{D_{kr} \}_{r = 0}^k$ contains $(i, j)$ and hence $(i, j)$ will be added to $E$ upon the end of the iteration.

  Now, still assuming that $i \in \pa{j}$, suppose instead that there is already an $i \rightarrow j$ path in $E$.  Using an induction hypothesis that $E$ contains edges if and only if they are in $\gcg$, then either $(i, j) \in E$ correctly, or there is more than one $i \rightarrow j$ path in $\gcg$, which is not possible since $\gcg$ is strongly causal.

  Finally, suppose that $i \not \in \pa{j}$.  If $i \not \pwgc j$ then we will correctly leave $(i, j)$ absent from $E$, so suppose that $i \pwgc j$.  Then we must have either $i \in \anc{j}$ or $\anc{i} \cap \anc{j} \ne \emptyset$; suppose first the former.  Then, there is some $u$ s.t. $i \pwgc u \pwgc j$, so by the construction of \hl{it seems} that both $(i, u)$ and $(u, j)$ should have already been added to $E$, so we have an $i \rightarrow j$ path in $E$ \hl{is this sound?}  Suppose then the latter case above, $u \in \anc{i} \cap \anc{j}$ \hl{try to construct a counterexample involving this case}.
\end{proof}

Proof of the actual theorem...

@@

% Given non-empty P and C...  Can we show all P -> C edges get collected in D?
  Firstly, if $n = 0$ or $n = 1$ the algorithm is trivially correct, returning on the first iteration with $E = \emptyset$.  Suppose then that $n > 1$ and let us define

  \begin{equation*}
    A_k = \{i \in [n]\ |\ \dist{i}{\widehat{P}_0} = k\},
  \end{equation*}

  where

  \begin{equation*}
    \dist{i}{P} = \text{max}\{r \in [n]\ |\ \exists (j \rightarrow i) \text{ path of length } r \text{ in } \gcg \text{ for some } j \in P\}.
  \end{equation*}

  We will proceed by induction and show that when the counter has reached the value $k \ge 1$, it is guaranteed that all nodes in $\bigcup_{\ell \le k}A_\ell$ (nodes having paths of length up to $k$ between themselves and $\widehat{P}_0$) must have all of their incident edges included in $E$, and that every edge in $E$ is an edge of $\gcg$.  If this is the case, then it can be immediately seen that the algorithm terminates with the correct result, $\gcg = ([n], E)$, once $k = n$; since there can only be paths up to length $n - 1$ in $\gcg$, and there cannot be any nodes or edges that don't involve paths back to $\widehat{P}_0$.  That is, $\bigcup_{\ell \le n}A_\ell = [n]$.

  % firstly dispensing with some trivialities, and then establishing that every edge in $\gcg$ makes an appearance in some $D_{kr}$ as well as that everything in any $D_{kr}$ is indeed an edge in $\gcg$.

  On the first iteration of the algorithm ($k = 0$) we have $S_0 = [n]$ and $E = \emptyset$.  To quickly dispense with some trivialities: the set $\widehat{P}_0$ is non-empty since $\gcg$ is a DAG and therefore there are nodes in $\gcg$ without parents, thence if $\widehat{P}_0 = \emptyset$ proposition \ref{prop:ancestor_properties} would be contradicted.  Now, if $\widehat{P}_0 = [n]$ then the graph necessarily contains no edges and we will correctly return $E = \emptyset$.

  Continuing with the non-trivial case: if $\widehat{P}_0 \ne [n]$ then $C_0$ must be non-empty, otherwise there would be driving nodes not included in $\widehat{P}_0$.  In this case it is also clear that $A_1 \ne \emptyset$, so let us consider some $j \in A_1$.  We must have $j \in C_0$, otherwise there would be a path of length $2$ from $\widehat{P}_0$ to $j$.  Moreover, since $\exists i \in \widehat{P}_0$ s.t. $i \gc j$ and therefore $i \pwgc j$ (by proposition \ref{prop:pwgc_anc}) we have $(i, j) \in D_{00}$ and therefore $(i, j) \in E$.  Therefore $A_1 \subseteq D_{00}$.  \hl{This is nonsensical because $A_1$ does not contain tuples -- need to show that $A_1 \subseteq C_0$ ?}

  Now consdider some $(i, j) \in D_{00}$ (the same reasoning as for $C_0$ is sufficient to see that $D_{00}$ is non-empty).  Since $i$ has no parents, $\anc{i} \cap \anc{j} = \emptyset$ and therefore $i \in \anc{j}$ (proposition \ref{prop:ancestor_properties}).  Furthermore, the construction of $C_k$ implies that $i \in \pa{j}$ since otherwise, there would be some $u \in \anc{j}$ with $i \in \anc{u}$, but then by proposition \ref{prop:pwgc_anc} $i \pwgc u$ and $u \pwgc j$ so that $u \not \in \widehat{P}_0 \implies u \in S_0 \implies j \not\in C_0$, a contradiction.  This implies every tuple in $D_{00}$ is a bona-fide edge of $\gcg$, as well as that $(i, j) \in A_1$.  So then $A_1 = D_{00}$, [\hl{nonsensical, see earlier.}] completing the base case for induction.

  Suppose now the algorithm has reached step $k$ and that $E = \bigcup_{\ell \le k - 1}A_\ell$.

% The reasoning here is (I think) correct -- but isn't directly establishing the induction hypothesis.
%   On the other hand, if $\widehat{P}_0 \ne [n]$ then $C_0$ must also be non-empty, otherwise there would be driving nodes not included in $\widehat{P}_0$.  Consider now an edge $(i, j) \in D_{00}$ ($D_{00}$ is non-empty for the same reason as $C_0$).  Since $i$ has no parents, $\anc{i} \cap \anc{j} = \emptyset$ and therefore $i \in \anc{j}$ (proposition \ref{prop:ancestor_properties}).  Furthermore, the construction of $C_k$ implies that $i \in \pa{j}$ since otherwise, there would be some $u \in \anc{j}$ with $i \in \anc{u}$, but then by proposition \ref{prop:pwgc_anc} $i \pwgc u$ and $u \pwgc j$ so that $u \not \in \widehat{P}_0 \implies u \in S_0 \implies j \not\in C_0$, a contradiction.  So every tuple in $D_{00}$ is a bona-fide edge of $\gcg$.  Suppose now that for some $j \in C_0$ there is an $i$ s.t. $(i, j) \in \gcg$ but $(i, j) \not\in D_{00}$.  Since $(i, j) \in \gcg$ we have $i \pwgc j$ (corollary \ref{cor:gc_implies_pwgc}), but then if $i \not\in \widehat{P}_0$ then $j \notin C_0$ and if $i \in \widehat{P}_0, j \in C_0$ then necessarily $(i, j) \in D_{00}$ since we cannot at this stage have any paths in $E$.
